---
title: "Statistical inference with the GSS data"
output: 
html_document: 
fig_height: 14
highlight: pygments
theme: spacelab
---
## Setup
```{r}
knitr::opts_chunk$set(echo = TRUE,fig.align='center', message = FALSE, warning=FALSE)

```


### Load packages

```{r load-packages, message = FALSE}
library(ggplot2)
library(dplyr)
library(statsr)
```

### Load data

```{r load-data}
load("./data/gss.Rdata")
```



* * *

## Part 1: Data

<b> I) DESCRIPTION </b>

The data used in this project comes from the <b>General Social Survey</b>, which is a
sociological survey collected by the University of Chicago.
It is one of the most distinguished studies in the social sciences and has
been monitoring the change in the society of the United States since 1972.
Some of the topics in the GSS are: marijuana use, crime, intergroup relations, social and economic life,
lifestyle, civil liberties, sexual behavior, etc.

Consequently, the General Social Survey provides social scientists &
statisticians with high-quality data, which can help them identify
trends & changes in american society.

<b> II) METHODOLOGY </b>

* The target population are adults (18+) living in households in the
United States.
* The sample is drawn using a proportional sampling technique,
using an area probability design that randomly selects respondents.
* Respondents are from a mix of urban, suburban, and rural geographic
areas.
* Participation in the study is voluntary.

<b>NOTE</b> that the General Social Survey is indeed, an <b><i>observational study.</i></b>

Therefore, statistical analyses using this data can only <b><i>identify association between independent variables and the outcome of interest.</i></b>


* * *

## Part 2: Research question
Research Question
Is there a considerable income inequality among families of different
races in the United States?

Using statistical inference methods, we'll assess whether
the data in the GSS supports the idea of a significant
inequality in the income of families from different races.

Remember that, since the GSS is an observational study, we can
reveal only association between the variables.
In this context, race might as well be a confounding variable,
and the difference in incomes may be caused by other social
factors such as achievement gaps, size of the family,
gender, etc.


* * *

## Part 3: Exploratory data analysis
## 3.1) Selecting the Data
* * *

For the Exploratory Data Analysis & the subsequent Inference, we'll
use the following variables:

* <b>coninc (income)</b>: Total family income in constant dollars (adjusted for inflation).
* <b>race</b>: Race of the respondent.
* <b>finrela</b>: Personal opinion of family income.


* * *
```{r Subsetting}
df <- gss %>% 
        select(race, coninc, finrela) %>%  
        na.omit() %>% 
        rename(income = coninc)
```
## Analyzing the Total Number of Survey Respondents
* * *

The very first thing we want to do is to get the the number
of people sampled by each race.
We can easily create a barplot & a table to evaluate this, which is what we'll do in the following code chunk.

* * *

```{r Number of Survey Respondents}
# Barchart
ggplot(data = df, mapping = aes(x = race, fill = race)) +
  geom_bar(aes(y = (..count..) / sum(..count..))) +
  labs(
    title = "Survey Respondents by Race",
    subtitle = "Bar Chart",
    caption = "Data: General Social Survey",
    x = "Race",
    y = "Percentage"
  ) +
  theme_light() +
  scale_fill_brewer(type = "qual", palette = 2) +
  theme(
    plot.title = element_text(face = "bold", hjust = .5),
    plot.subtitle = element_text(
      face = "bold",
      hjust = .5,
      color = "dimgrey"
    ),
    plot.caption = element_text(color = "gray"),
    axis.title.x = element_text(face = "bold",
                                color = "dimgrey"),
    axis.title.y = element_text(face = "bold",
                                color = "dimgrey"),
  ) + guides(col = "none", fill = "none")

# Respondents Count & Percentage grouped by Race
df %>% 
        group_by(race) %>% 
        summarize(n_respondents = n()) %>% 
        mutate(p_respondents = n_respondents/sum(n_respondents))

```
* * *

We've discovered that approximately 81.2% of the people sampled in this data are white, which is a huge difference with 
respect to the other races.
Nevertheless, other sample sizes are also large, with almost 8K black people sampled & 2.7K from other races.


* * *
## Opinion of Family Income
* * *
One of the most interesting variables in this data set is the finrela
variable, which categorizes the opinion of the respondent
with respect to their family's income. 
Using this particular variable, we will analyze whether the income perception differs between races or not.

The following code chunk creates a dataframe, where the percent 
variable is the relative frequency with respect to each race.
(Hence, the percent of each race sums to 1).


<b>NOTE:</b> We compare the relative frequencies to adjust for the large difference in sample sizes we found in the last bar chart.
* * *
```{r Table: Opinion of Family Income}


data <- 
        df %>% 
        mutate(
                finrela = case_when(
                        finrela == "Far Below Average" ~ "Below Average",
                        finrela == "Far Above Average" ~ "Above Average",
                        TRUE ~ as.character(finrela)
                ),
                finrela = factor(finrela, ordered = T, 
                                 levels = c("Below Average", "Average",
                                            "Above Average"))
        ) %>% 
        group_by(race, finrela) %>% 
        summarise(n=n()) %>%
        transmute(finrela, percent = n/sum(n))

data
```
* * *

Now we can create a barplot using the ggplot2 package in order
to observe the opinion of family income grouped by race.

* * *

```{r Plot: Opinion of Family Income}

ggplot(data = data,
       mapping = aes(x = finrela,
                     y = percent,
                     fill = race)) +
        geom_bar(stat = "identity") +
        labs(title = "Opinion of Family Income",
             x = "",
             y = "Count") +
        facet_wrap(. ~ race) +
        theme_light() +
        theme(
                plot.title = element_text(face = "bold", hjust = .5),
                axis.text.x = element_text(
                        face = "bold",
                        color = "dimgrey",
                        angle = 90
                ),
                strip.background = element_rect(fill = "red2"),
                strip.text = element_text(face = "bold", color = "white")
        ) + guides(fill = "none")


```
* * *

We can observe that the average category is the most frequent one
in every single panel.
Also, the barplot that represents white people is fairly 
symmetric.
This is not the case with black and other races (the barplot is
left-skewed), which clearly implies they have a rather negative view of their family's income.
In summary, we have found that black & other races tend to view themselves as having a below average income, more often than whites.
Also, white people tend to view themselves as having an above average
income, more often than other races.

This is quite an intriguing result.
We'll see if it's justified later on...

* * *

## Distribution of the Total Family Income
```{r Distribution of Total Family Income}
# Histogram of the Income variable
ggplot(data = df, mapping = aes(x = income)) +
        geom_histogram(fill = "skyblue", col = "black") +
        theme_minimal() +
        labs(
                title = "Total Family Income",
                subtitle = "Histogram",
                caption = "Data: General Social Survey",
                x = "Income (Constant U$D)",
                y = "Count"
        ) +
        theme_classic() +
        theme(
                plot.title = element_text(face = "bold", hjust = .5),
                plot.subtitle = element_text(face = "bold", hjust = .5,
                                             color = "dimgrey"),
                plot.caption = element_text(color = "gray"),
                axis.title.x = element_text(face = "bold",
                                            color = "dimgrey")
        )

# Summary of the Total Family Income
with(df, summary(income))
```

# Variability & Mean of Family Income
```{r Variability of Total Family Income}
# Violin Plot: Income by Race
ggplot(data = df, mapping = aes(x = income, y = race, fill = race)) +
        geom_violin(alpha = .9, color = "gray") +
        geom_boxplot(width=0.1, color = "black") +
        stat_summary(fun = mean, col = "red", aes(shape = "Mean")) +
        labs(
                title = "Income by Race",
                subtitle = "Violin Plot with Boxplot",
                caption = "Data: General Social Survey",
                x = "Income",
                y = "Race"
        ) +
        scale_fill_brewer(palette = "Set3") +
        scale_color_viridis_d(option = "viridis", begin = .7) +
        theme(
                plot.title = element_text(face = "bold", hjust = .5),
                plot.subtitle = element_text(face = "bold", hjust = .5,
                                             color = "dimgrey"),
                plot.caption = element_text(color = "gray"),
                axis.title.x = element_text(face = "bold",
                                            color = "dimgrey"),
                axis.title.y = element_text(face = "bold",
                                            color = "dimgrey"),
                legend.title = element_text(face = "bold",
                                            color = "dimgrey")
        ) + guides(col = "none", fill = "none", 
                   shape = guide_legend(title = "Statistic"))

# --- Quantiles, shown in the Boxplot ---
df %>% 
        na.omit() %>% 
        group_by(race) %>% 
        do(data.frame(t(quantile(.$income, probs = c(.25, .5, .75)))))

# --- Income Mean by Race (Red Dot) ---
# Forcing Linear Regression through the Origin and
# getting it's coefficients
lm(data = df, formula = income ~ race - 1) %>% coef()

```


* * *

## Part 4: Inference

NOTE: Insert code chunks as needed by clicking on the "Insert a new code chunk" 
button (green button with orange arrow) above. Make sure that your code is visible
in the project you submit. Delete this note when before you submit your work.